---
title: "Rare Diseases"
date: "`r Sys.Date()`"
author: Jenny Sloane
output:
  rmdformats::readthedown:
    self_contained: true
---

<style type="text/css">

body {
  font-size: 12pt;
}

pre {
  font-size: 17px;
}

h1.title, h1.title p {
  font-size: 40px;
}

h1 { 
  font-size: 35px;
}

</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

# Orphadata

- <a href="https://www.orphadata.com/" target="_blank">Link to Orphadata website</a>
- <a href="https://www.orphadata.com/" target="_blank">Link to data</a>
- **Rare Disease** = A disease that affects fewer than 200,000 people in the United States
- US population = 331.9 million 
- 200,000/331.9 million = ~0.0006
- 6-9/10,000 = .0006 - .0009
<!-- - 1-5/10,000 = .0001 - .0005 -->

```{r}
library(xml2)
library(tidyverse)
library(ggeasy)
require(httr)
library(janitor)
library(gt)

my_url <- "https://www.orphadata.com/data/xml/en_product9_prev.xml"
disease_xml <- as_list(read_xml(my_url))

raw_data <- disease_xml$JDBOR

data_clean = tibble::as_tibble(raw_data) %>%
  unnest_wider(DisorderList) %>% 
  unnest(Name, OrphaCode, ExpertLink, DisorderType, DisorderGroup) %>% 
  unnest(DisorderType, DisorderGroup) %>% 
  unnest(PrevalenceList) %>% 
  unnest_wider(PrevalenceList) %>% 
  unnest(PrevalenceClass, PrevalenceType, PrevalenceQualification, PrevalenceGeographic, PrevalenceValidationStatus, ValMoy,Source) %>% 
  unnest(PrevalenceClass, PrevalenceType, PrevalenceQualification, PrevalenceGeographic, PrevalenceValidationStatus) %>% 
  select(-Availability) %>% 
  mutate_all(as.character) %>%
  clean_names()

```

# Graphing the distribution of diseases 

- Note: some diseases are counted multiple times (when there are multiple sources for the same disease)
  - Example: Cleft palate is counted 10 times because there's data from different geographic regions 

```{r fig.width=10, fig.height=7}
bar_plot <- data_clean %>%
  filter(!(prevalence_class %in% c("Unknown", "NULL", "Not yet documented"))) 

bar_plot$prevalence_class <- factor(bar_plot$prevalence_class, levels = c("<1 / 1 000 000", "1-9 / 1 000 000", "1-9 / 100 000", "1-5 / 10 000", "6-9 / 10 000", ">1 / 1000"))

ggplot(bar_plot,aes(x=prevalence_class, fill=factor(ifelse(prevalence_class=="6-9 / 10 000","Highlighted","Normal")))) +
  geom_bar() +
  scale_fill_manual(name = "prevalence_class", values=c("red","grey50")) +
  stat_count(geom = "text", colour = "black", size = 5,
aes(label = ..count..),position=position_stack(vjust=1)) +
  labs(x = "Prevalence", y = "Frequency", title = "Distribution of Prevalence of Rare Diseases") +
  theme_bw() +
  easy_remove_legend() +
  easy_text_size(20) +
  theme(axis.text = element_text(size = 15)) 

```


# Disorder types

- 6-9 / 10,000 only

```{r}
rd_names <- data_clean %>%
  filter(prevalence_class == "6-9 / 10 000") %>%
  select(orpha_code, name, disorder_type, prevalence_class, val_moy, prevalence_geographic, prevalence_validation_status)

rd_names %>%
  group_by(disorder_type) %>%
  count() %>%
  arrange(desc(n)) 

```

# 6-9 / 10,000 "Disease"

```{r}
library(reactable)

disease_names <- rd_names %>%
  filter(disorder_type == "Disease")

# unique(disease_names$name)

rd_names %>%
  filter(disorder_type == "Disease") %>%
  select(name, prevalence_geographic) %>%
  reactable(striped = TRUE)

```

# ALS 

- average of 9-12 months to diagnosis
- <a href="https://www.orpha.net/consor/cgi-bin/Disease_Search.php?lng=EN&data_id=106&Disease_Disease_Search_diseaseGroup=Amyotrophic-lateral-sclerosis&Disease_Disease_Search_diseaseType=Pat&Disease(s)/group%20of%20diseases=Amyotrophic-lateral-sclerosis&title=Amyotrophic%20lateral%20sclerosis&search=Disease_Search_Simple" target="_blank">Link to Orpha Net: ALS</a>

```{r}
data_clean %>%
  filter(grepl('ALS|Amyotrophic|amyotrophic', name)) %>%
  select(name, prevalence_class, prevalence_geographic) %>%
  reactable(striped = TRUE)
```

# Myositis
- average of over 2 years to diagnosis

```{r}
data_clean %>%
  filter(grepl('myositis|Myositis', name)) %>%
  select(name, prevalence_class, prevalence_geographic) %>%
  reactable(striped = TRUE)
```

# Vasculitis 
- average of 7 months to diagnosis

```{r}
data_clean %>%
  filter(grepl('vasculitis|Vasculitis', name)) %>%
  select(name, prevalence_class, prevalence_geographic) %>%
  reactable(striped = TRUE)
```

